\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-09 paper/ms.tex
Some edits to allay referee's concerns

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales as the cube of the number of data points,
their application has been limited to relatively small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semiseparable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in some physical systems and we demonstrate it can be used as
an effective model in some other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the cost of
computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
This means that for large datasets, every evaluation of the likelihood quickly
becomes computationally intractable, and representing the matrix in memory can
become expensive.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

@@

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

\section{Summary}

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in many cases, small values of $J$ are sufficient.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to model stellar oscillations using the relation to the
mixture of stochastically-driven, damped simple harmonic oscillators.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

@@

@@

To help clarify equations \ref{system}, we write out here an example of a semiseparable matrix with a
single \celeriteterm\ term with $c=d=1$:
\begin{eqnarray}\eqlabel{semi-separable-structure}
K&=&%
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&=&
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost commonly scales as the cube of the number of data points,
their application has been limited to relatively small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semiseparable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in physical systems and we demonstrate it can be used as
an effective model in other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets, $N \approx 10^3$, because the cost of
computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
This means that for large datasets, every evaluation of the likelihood quickly
becomes computationally intractable, and representing the matrix in memory can
become expensive.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

@@

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero_psd}}
\end{center}
\end{figure}

@@

\section{Summary and discussion}

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost generally
scales as the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in some cases, small values of $J$ may suffice.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to model stellar oscillations using the relation to the
mixture of stochastically-driven, damped simple harmonic oscillators, as
we have demonstrated in this paper.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

\subsection{Other applications and limitations}

@@

\subsection{Code availability}

@@

To help clarify equations \eq{system}, we write out here an example of a semiseparable matrix with a
single \celeriteterm\ term with $c=d=1$:
\begin{eqnarray}\eqlabel{semi-separable-structure}
K&=&%
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&=&
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,2J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.
